---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

For reading the file we need to process two steps:  
1. Unzip the archive to CSV-file  
2. Read the CSV-file

Next R-code realize these steps. To control if result is correct, print result dimensions.

```{r read file}
unzip_file <- unzip("activity.zip")
activity <- read.csv(unzip_file)
dim(activity)
```

## What is mean total number of steps taken per day?

For this part of the assignment, the missing values in the dataset are ignored.

*activity$date* is a factor, so for counting total number of steps per day *tapply*
can be used. Result will be put to a *steps_total* variable.

```{r total number of steps}
steps_total <- tapply(activity$steps, activity$date, sum)
hist(steps_total, xlab = "Steps per day")
```

Now when we have total numbers of steps per day for every date in the dataset, we
can count mean and median numbers (using r functions) of steps per day.

Mean total number of steps per day:
```{r mean numbers of steps}
steps_mean <- mean(steps_total, na.rm = T)
steps_mean
```

Median total number of steps per day:
```{r median numbers of steps}
steps_median <- median(steps_total, na.rm = T)
steps_median
```

## What is the average daily activity pattern?

In original dataset *activity\$interval* is a integer vector. In this case we will need to
us *as.factor* function to use *split* to subset original dataset by intervals.

```{r split steps by interval}
steps_by_interval <- split(activity, as.factor(activity$interval))
intervals <- names(steps_by_interval)
head(intervals)
```

Now we can count *mean* and *median* of steps in every subset by interval. NA will be ignored.
```{r total split steps by interval}
steps_average_per_interval <- data.frame(interval = rep(0, length(intervals)),
                                         steps = rep(0, length(intervals)),
                                         median_steps = rep(0, length(intervals)))
for (i in 1:length(intervals)){
    steps_average_per_interval[i, 1] <- intervals[i]
    steps_average_per_interval[i, 2] <- mean(steps_by_interval[[i]][['steps']], na.rm = T)
    steps_average_per_interval[i, 3] <- median(steps_by_interval[[i]][['steps']], na.rm = T)
}

plot(steps_average_per_interval$interval, steps_average_per_interval$steps,
     type = 'l', xlab = "Intervals", ylab = "Average number of steps", col = "blue")
```

On this point we have a data frame that in first column has all intervals and in second
column has corresponding average number of steps, so we can find an interval that in average
has the maxumum number of steps.
```{r interval with maximum steps}
max_index <- which.max(steps_average_per_interval$steps)
max_interval <- steps_average_per_interval[max_index, 1]
max_interval
```

## Imputing missing values

Calculate number of *NA* in *activity* by columns

```{r number of NA in dataset}
# NA number in "date"
na_number <- sum(is.na(activity$date))
na_number
# NA number in "intervals"
na_number <- sum(is.na(activity$interval))
na_number
# NA number in "steps"
na_number <- sum(is.na(activity$steps))
na_number
```
As we can see, all missing values are situated in "steps" column.

Like a strategy for replacement of messing values was chosen the next one: every
missing value will be replaced with median for corresponding time interval. This strategy
lets us to ignore some abnormal person behavior.

For the first copy *activity* data frame to a new one and select *steps* coulumn to
some variable for the future usage.
```{r create new data frame}
new_activity <- activity
steps <- new_activity$steps
```

Now we can replace every *NA* with corresponding median for interval, that we calculated earlier.
```{r replace NA with median}
for (i in 1:length(steps)) {
    if (is.na(steps[i])) {
        interval <- new_activity[[i, 3]]
        median_steps <- steps_average_per_interval[steps_average_per_interval$interval == interval, 3]
        steps[i] <- median_steps
    }
}
new_activity$steps <- steps
```
## Are there differences in activity patterns between weekdays and weekends?
